Assessing primitives performance\\ on multi-stage execution

### 2018-05-08 paper.tex
update plots, micro bench parts + start asm explanation

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{ifthen}
\usepackage{pgfplots}
\usepgfplotslibrary{statistics}
%\input{macros.tex}

\usepackage{xcolor}
\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sk}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Sophie:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}

@@

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to improve the execution of the client language with a JIT performing speculative optimizations \cite{ursPHD} for the execution of the optional primitives implemented in the client language to be as fast as the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were to first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations, in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve,
	\item The performance of the code not yet optimized is much worse than the peak performance code,
	\item Even mature JIT fail to optimize some narrow cases, where the performance is drastically slower.
\end{itemize}
Overall, this approach requires high engineering time to get a mature optimizing JIT with speculative optimizations and even then, it leads in practice to unreliable performance.

VMs are traditionally implemented in a low-level language such as C or C++. To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippets are done through an interpreter and the JIT compiles at runtime the frequently used code snippets for further executions to be performed quicker. This leads to annoying concerns. For example, let's say we implement a primitive in C. Frequently used portion of code calling that primitive are going to be compiled to native code by the JIT. This is problematic since the native code generated by the JIT is not directly compatible with native code generated by the C compiler: a call from one to another may require to edit the stack pointer and the frame pointer from the client stack to the C stack and to spill or move multiple registers. Switching between both can waste up to around a dozen native instructions. If the primitive is going to execute many instructions, the switch overhead might be negligible, but if the primitive executes only a few instructions, the overhead can be noticeable.

In the paper, we took the example of a string comparison optional primitive. The primitive compares two strings and answers if one string is greater, equal or lower than the other string. The exact specification are detailed in Section \ref{subsec:primSpec}. We implemented a first version in Smalltalk. We then implemented multiple version compiling through C compilers to native code, using different part of our VM infrastructure. Lastly, we implemented a version in the back-end of the JIT. Each time, we tried to write the code in the most efficient way possible. Then, we compared the execution time of all versions, showing that overall the most complex implementation is the fastest to execute.

@@

Because of this design, switching the execution between the interpreter and the native code generated by the JIT has a cost. Effectively, the VM needs to switch the frame pointer, the stack pointer and depending on architecture the link register from one stack to the other one. It also needs to save those pointers to be able to resume execution from one stack to the other one. In addition, caller-saved registers on the C conventions or the JIT convention need to be saved since the VM cannot tell ahead what code is going to be used in the other execution model and what registers are required. We estimate the overhead to up to around a dozen of processor instruction, the exact number varies a lot so it's difficult to precise it (the number of register to edit depends on the processor and on the code executed).

Most of the primitives are implemented either in Slang or directly in C. They benefit from the C compiler optimizations to be efficient. When called from the interpreter, activating a primitive is just a C call (the primitive function pointer is cached in the look-up cache next to the method to activate). When called from the native code generated by the JIT, activating a primitive requires to switch the execution form the client stack to the C stack, perform the primitive, and switch back to resume execution. If the primitive takes a significant amount of instructions to execute, for example, the primitive is copying 100kb from one array to another one, the stack switch dance overhead is negligible. If the primitive is very quick to perform, for example, the primitive is adding two integers which do not overflow, the overhead is massive: the execution cost of the addition goes from a few processor instructions to a few dozen instructions.

@@

 \cba{petite intro avec plan de section, context was moved before.}

 \subsection{Different implementation and execution}

@@

\paragraph{Baseline: pure Smalltalk.}
: baseline
Smalltalk code on top of VM. Used as our baseline for benchmarks.
Write down the Smalltalk code here.
%CHECK how to import code env in latex.
%Note we're talking about comparison, not =, so not sure we should show = or bench =
%\begin{code}
%= aString 	
%	self == aString ifTrue: [ ^true ].
%	aString isString ifFalse: [ ^false ].
%	self size = aString size ifFalse: [ ^false ].
%	1 to: self size do: [:i | (self at: i) = (aString at: i) ifFalse: [ ^ false ] ].
%	^ true
%\end{code}
Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:

New spec

@@

@@

\subsection{Micro-benchmarks}
Results on strings of different lengths (3, 10, 1000)

@@

    \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		xbar width=0.30,
		no markers,Baseline/.style={red},2/.style={green},3/.style={blue}
		%cycle list={red, orange, green, purple, black}
        ]
\addplot[
    fill=black!25,
    %draw=none,
    %fill,
    draw=black,
    fill,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
]  
table [y error=error] {
x   y           error    
1   1           0 
2   0.3147064   0.002288488 
3   0.3035965   0.001786185
4   0.8545431   0.006132950 
5   2.250732    0.04133638 
};            
      \end{axis}
    \end{tikzpicture}
    \subcaption{Size 0}
    \end{subfigure}
    \begin{subfigure}[b]{.45\textwidth}
    \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   1.2505193   0.004611354 
3   1.2295801   0.001810171
4   2.8205415   0.008035024 
5   7.089806    0.01403785 
};            
      \end{axis}
    \end{tikzpicture}

    \subcaption{Size 5}
    \end{subfigure}
    \begin{subfigure}[b]{.45\textwidth}
 \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		ymin=0,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   6.6951738   0.062391122 
3   6.6032572   0.024774416
4   10.3029394   0.061481532 
5   17.979611    0.10127936 
};            
      \end{axis}
    \end{tikzpicture}

@@

     \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		ymin=0,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   9.4064534   0.036698187 
3   9.2138415   0.013982843
4   13.4384830   0.040202717 
5   26.404792    0.09298324 
};            
      \end{axis}
    \end{tikzpicture}

   \subcaption{Size 1000}
   \end{subfigure}
   \caption{GCC-Linux results}
\end{figure*}

@@

    \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		ymin=0,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   0.3147064   0.002288488 
3   0.3035965   0.001786185
4   0.8545431   0.006132950 
5   2.250732    0.04133638 
};            
      \end{axis}
    \end{tikzpicture}

@@

    \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   1.2505193   0.004611354 
3   1.2295801   0.001810171
4   2.8205415   0.008035024 
5   7.089806    0.01403785 
};            
      \end{axis}
    \end{tikzpicture}
    \subcaption{Size 5}
    \end{subfigure}
    \begin{subfigure}[b]{.45\textwidth}
 \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		ymin=0,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   6.6951738   0.062391122 
3   6.6032572   0.024774416
4   10.3029394  0.061481532 
5   17.979611   0.10127936 
};            
      \end{axis}
    \end{tikzpicture}
    \subcaption{Size 100}
    \end{subfigure}
    \begin{subfigure}[b]{.45\textwidth}

    \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		ymin=0,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   9.4064534   0.036698187 
3   9.2138415   0.013982843
4   13.4384830  0.040202717 
5   26.404792   0.09298324 
};            
      \end{axis}
    \end{tikzpicture}

   \subcaption{Size 1000}
   \end{subfigure}
   \caption{LLVM-Mac results}
\end{figure*}

\cba{Je pense string de taille 0-1 c'est cool pour montrer que le code C va moins vite que pur smalltalk dans ce cas (si c'est vraiment le cas)}
4 string length. (1er <=3, dernier > 1000) 
line 1: GCC
line 2: LLVM
String evaluation here.

Il faut expliquer en une phrase: 
petit string = cout important C-St stack switch.
Gde string = cout de la boucle.

>>>>

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{ifthen}
\usepackage{pgfplots}
\usepackage{listings}
\usepgfplotslibrary{statistics}
%\input{macros.tex}

\usepackage{xcolor}
\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sk}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Sophie:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand*\rot{\rotatebox{90}}
\input{macros}

@@

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to improve the execution of the client language with a JIT performing speculative optimizations \cite{ursPHD} for the execution of the optional primitives implemented in the client language to be as fast as the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were \sk{the?} to first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve,
	\item The performance of the code not yet optimized is much worse than the peak performance code,
	\item Even mature JIT fail to optimize some narrow cases, where the performance is drastically slower.
\end{itemize}
Overall, this approach requires high engineering time to get a mature optimizing JIT with speculative optimizations and even then, it leads in practice to unreliable performance.

VMs are traditionally implemented in a low-level language such as C or C++. To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippet are done through an interpreter and the JIT compiler at runtime the frequently used code snippets for further executions to be performed quicker. This leads to annoying concerns: for example, let's say we implement a primitive in C. Frequently used portion of code calling that primitive are going to be compiled to native code by the JIT. This is problematic since the native code generated by the JIT is not directly compatible with native code generated by the C compiler: a call from one to another may require to edit the stack pointer and the frame pointer from the client stack to the C stack and to spill or move multiple registers. Switching between both can waste up to around a dozen native instructions. If the primitive is going to execute many instructions, the switch overhead might be negligible, but if the primitive executes only a few instructions, the overhead can be noticeable.

In this paper, we take the example of a string comparison optional primitive. The primitive compares two strings and answers if one string is greater, equal or lower than the other string. The exact specifications are detailed in Section \ref{subsec:primSpec}. We implemented a first version in Smalltalk. We then implemented multiple version compiling through C compilers to native code, using different part of our VM infrastructure. Lastly, we implemented a version in the back-end of the JIT. Each time, we tried to write the code in the most efficient way possible. Then, we compared the execution time of all versions, showing that overall the most complex implementation is the fastest to execute.

@@

Because of this design, switching the execution between the interpreter and the native code generated by the JIT has a cost. Effectively, the VM needs to switch the frame pointer, the stack pointer and depending on architecture the link register from one stack to the other one. It also needs to save those pointers to be able to resume execution from one stack to the other one. In addition, caller-saved registers on the C conventions or the JIT convention need to be saved since the VM cannot tell ahead what code is going to be used in the other execution model and what registers are required. We estimate the overhead to up to around a dozen of processor instruction, the exact number varies a lot so it is difficult to precise it (the number of register to edit depends on the processor and on the code executed).

Most of the primitives are implemented either in Slang or directly in C. They benefit from the C compiler optimizations to be efficient. When called from the interpreter, activating a primitive is just a C call (the primitive function pointer is cached in the look-up cache next to the method to activate). When called from the native code generated by the JIT, activating a primitive requires to switch the execution from the client stack to the C stack, perform the primitive, and switch back to resume execution. If the primitive takes a significant amount of instructions to execute, for example, the primitive is copying 100kb from one array to another one, the stack switch dance overhead is negligible. If the primitive is very quick to perform, for example, the primitive is adding two integers which do not overflow, the overhead is massive: the execution cost of the addition goes from a few processor instructions to a few dozen instructions.

@@

 This section first describes the different ways of implementing a primitive. We then explain how each of these version are executed by the VM and the pros and cons of each and every.

 \subsection{Different implementations and executions}

@@

\paragraph{Baseline: pure Smalltalk.}
: baseline
Smalltalk code on top of VM. Used as our baseline for benchmarks.

\begin{code}
compareWithPureST: aString

	| c1 c2 length1 length2 |
	length1 := self size.
	length2 := aString size.
		1 to: (length1 min: length2) do: [ :i |
			(c1 := self basicAt: i) = (c2 := aString basicAt: i) 
			ifFalse: [
				c1 < c2
					ifTrue: [ ^1 ]
					ifFalse: [ ^3 ] ] ].
		length1 = length2 ifTrue: [ ^2 ].
		length1 < length2
			ifTrue: [ ^1 ]
			ifFalse: [ ^3 ] 
\end{code}

Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:
New spec

@@

\begin{table}
\centering
\begin{tabular}{|c||c|c|c|c|}
  \hline
    & \rot{Engineering time} & \rot{Execution time} & \rot{Readability}  \\
  \hline
  Baseline & + & +++ & +++ \\
  SmartSyntax & ++ & ++ & ++ \\
  SlangPlugin & ++ & ++ & ++ \\
  PureSlang & +++ & + & ++ \\
  Slang+JIT & +++ & + & + \\
  \hline
\end{tabular}
\caption{Trade-offs between implementations: synthetized}
\label{tab:comparisontable}
\end{table}

Each of these implementations has its pros and cons considering the context. In the end, there is always a trade-off between engineering time and performance. In

@@

\subsection{Micro-benchmarks}

@@

    \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		xbar width=0.30,
		no markers,Baseline/.style={red},2/.style={green},3/.style={blue}
		%cycle list={red, orange, green, purple, black}
        ]
\addplot[
    fill=black!25,
    %draw=none,
    %fill,
    draw=black,
    fill,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
]  
table [y error=error] {
x   y           error    
1   1           0 
2   0.3411212   0.0006746299 
3   0.3375418   0.001577878
4   1.077762   0.005363974 
5   2.490423    0.004304882 
};            
      \end{axis}
    \end{tikzpicture}
    \subcaption{Size 0}
    \end{subfigure}
    \begin{subfigure}[b]{.45\textwidth}
    \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   1.3891212   0.0029489731 
3   1.3149840   0.020471694
4   3.818961   0.054523074 
5   6.740474    0.010805042 
};            
      \end{axis}
    \end{tikzpicture}

    \subcaption{Size 5}
    \end{subfigure}
    \begin{subfigure}[b]{.45\textwidth}
 \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		ymin=0,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   8.6795487   0.0586762222 
3   8.9585208   0.116958804
4   15.809515   0.300447885 
5   16.752667    0.106255682 
};            
      \end{axis}
    \end{tikzpicture}

@@

     \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		ymin=0,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   13.6051267   0.0589164968 
3   14.7789145   0.147140189
4   24.417751   0.186326223 
5   24.480071    0.095038374 
};            
      \end{axis}
    \end{tikzpicture}

   \subcaption{Size 1000}
   \end{subfigure}
   \caption{GCC-Linux results}
   \label{fig:gcc}
\end{figure*}

@@

    \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		ymin=0,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   0.3147064   0.002288488 
3   0.3035965   0.001786185
4   0.8545431   0.006132950 
5   2.250732    0.04133638 
};            
      \end{axis}
    \end{tikzpicture}

@@

    \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   1.2505193   0.004611354 
3   1.2295801   0.001810171
4   2.8205415   0.008035024 
5   7.089806    0.01403785 
};            
      \end{axis}
    \end{tikzpicture}
    \subcaption{Size 5}
    \end{subfigure}
    \begin{subfigure}[b]{.45\textwidth}
 \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		ymin=0,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   6.6951738   0.062391122 
3   6.6032572   0.024774416
4   10.3029394  0.061481532 
5   17.979611   0.10127936 
};            
      \end{axis}
    \end{tikzpicture}
    \subcaption{Size 100}
    \end{subfigure}
    \begin{subfigure}[b]{.45\textwidth}

    \begin{tikzpicture}
      \begin{axis}
      [
        xtick={1,...,5},
        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
        %x tick label as interval,
        font={\tiny},
		ybar=10pt,
		ymin=0,
		xbar width=0.30
        ]
\addplot[
    fill=black!25,
    draw=black,
    point meta=y,
    every node near coord/.style={inner ysep=5pt},
    error bars/.cd,
        y dir=both,
        y explicit
] 
table [y error=error] {
x   y           error    
1   1           0 
2   9.4064534   0.036698187 
3   9.2138415   0.013982843
4   13.4384830  0.040202717 
5   26.404792   0.09298324 
};            
      \end{axis}
    \end{tikzpicture}

   \subcaption{Size 1000}
   \end{subfigure}
   \caption{LLVM-Mac results}
   \label{fig:llvm}
\end{figure*}

%\cba{Je pense string de taille 0-1 c'est cool pour montrer que le code C va moins vite que pur smalltalk dans ce cas (si c'est vraiment le cas)}
We performed the benchmarks upon 4 different string sizes: 0, 5, 100 and 1000 characters, using 2 different compilers: GCC on one hand and LLVM on the other hand.

The results are shown in terms of relative speed-up towards the baseline implementation, i.e. the pure smalltalk version. 
We can notice some patterns over the results in figures \ref{fig:gcc} and \ref{fig:llvm}:
\begin{itemize}
\item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than the baseline's for the small strings. This can be explained by the high cost of the switching between the C and Smalltalk stacks, as mentionned in \sk{faire ref à la partie}.
\item The Slang+JIT implementation is more performant in every case. Nevertheless, 
the performance gap between this implementation and the PureSlang one tends to decrease as the string size increase. The comparison loop \sk{ref to appendix}
\end{itemize} proves to be more and more time-consuming because of the higher number of iterations.

Howerver, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{gcc} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factor for a 1000-characters string. They still greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factor.

%\begin{lstlisting}[language={[x64]Assembler}]
%movzbl 0x8(%eax,%edx,1),%ebp
%movzbl 0x8(%eax,%ebx,1),%ecx
%cmp    %ecx,%ebp
%jne    6e8 <primitiveCompareWith+0x1f8>
%add    $0x1,%eax
%cmp    %eax,%edi
%jne    610 <primitiveCompareWith+0x120> 
%\end{lstlisting}

\begin{table}
\centering
\begin{tabular}{|r|l|}
  \hline
   GCC & LLVM  \\
  \hline
  `movzbl 0x8(\%eax,\%edx,1),\%ebp` & `movl -0x14(\%ebp), \%esi`  \\
  movzbl 0x8(\%eax,\%ebx,1),\%ecx & movzbl 0x8(\%edx,\%esi), \%esi \\
  cmp    \%ecx,\%ebp & movl -0x18(\%ebp), \%edi \\
  jne    0x8558 &  movzbl 0x8(\%edx,\%edi), \%edi \\
  add    \$0x1,\%eax & subl \%edi, \%esi \\
  cmp    \%eax,\%edi &  jne 0x1f660 \\
  jne    0x8480 & incl \%edx \\
   & cmpl \%ebx, \%edx \\
   & jl 0x1f644 \\
  \hline
\end{tabular}
\caption{Different compilers optimizations: GCC and LLVM}
\label{tab:cmp_asm}
\end{table}
